---
title: "Dynamic Factor Analysis ensemble clustering for `r params$stock`"
date: "`r Sys.Date()`"
output: html_document
params:
   stock: "Haddock"
   timeseries_data: !r clusterthat::haddock_ts
   model_data: !r clusterthat::haddock_mod
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache.path = paste0(params$stock, "-cache/"),
  fig.path = paste0(params$stock, "-fig/")
)
```

```{r setup}
library(clusterthat)
library(ggplot2)
library(dplyr)
library(bayesdfa)
theme_set(theme_minimal())
options(mc.cores = parallel::detectCores())
```

# An example of clustering with Bayesian Dynamic Factor Analysis

Dynamic Factor Analysis (DFA) is a timeseries dimension reduction technique that estimates a shared set of trends and loading factors on those trends for the various timeseries. bayesdfa fits Bayesian DFA via a model written in Stan. Here we will demonstrate an example with the bayesdfa package.

<https://CRAN.R-project.org/package=bayesdfa>
<http://rjournal.github.io/archive/2019/RJ-2019-007/index.html>

```{r dfa-ts}
ggplot(params$timeseries_data) + geom_line(aes(
  x = year, y = fbar,
  colour = model_id, group = model_id
)) +
  scale_colour_viridis_c()

ggplot(params$timeseries_data) + geom_line(aes(
  x = year, y = ssb,
  colour = model_id, group = model_id
)) +
  scale_colour_viridis_c()
```

First we need to reshape our data for fitting:

```{r}
df <- params$timeseries_data %>%
  reshape2::dcast(year ~ model_id, value.var = "fbar") %>%
  dplyr::select(-year)

df_ssb <- params$timeseries_data %>%
  reshape2::dcast(year ~ model_id, value.var = "ssb") %>%
  dplyr::select(-year)
```

The columns are our various models and the rows represent years.

We will fit a model with 1 and 2 underlying trends for speed.

```{r dfa-fitting, warning=FALSE, results="hide", cache=TRUE}
m_fbar <- find_dfa_trends(y = df, kmin = 1, kmax = 2, iter = 150,
  convergence_threshold = 1.2, variance = c("equal"), data_shape = "long")

m_ssb <- find_dfa_trends(y = df_ssb, kmin = 1, kmax = 2, iter = 150,
  convergence_threshold = 1.2, variance = c("equal"), data_shape = "long")
```

Now compare the leave-one-out information criteria:

```{r}
# loo(m_fbar1)
# loo(m_fbar2)
# 
# loo(m_ssb1)
# loo(m_ssb2)

m_fbar_best <- m_fbar$best_model
m_ssb_best <- m_ssb$best_model
```

We will apply a varimax rotation to describe as much of the variation in the initial trend and then plot the underlying trends:

```{r}
rot_fbar <- rotate_trends(m_fbar_best)
plot_trends(rot_fbar)

rot_ssb <- rotate_trends(m_ssb_best)
plot_trends(rot_ssb)
```

We can also plot the loadings:

```{r, warnings = FALSE}
g_fbar <- plot_loadings(rot_fbar)
# remove colour if desired:
g_fbar$layers[[1]] <- NULL
g_fbar + geom_violin(colour = NA, fill = "grey30")

g_ssb <- plot_loadings(rot_ssb)
# remove colour if desired:
g_ssb$layers[[1]] <- NULL
g_ssb + geom_violin(colour = NA, fill = "grey30")
```

We can now use the loadings in a clustering analysis.

The following are our rotated loadings. We will reshape them into a data frame.

```{r}
dfa_loadings_fbar <- rot_fbar$Z_rot_mean %>%
  reshape2::melt() %>%
  rename(time = Var1, model_id = Var2, loading = value) %>%
  reshape2::dcast(time ~ model_id, value.var = "loading")

dfa_loadings_ssb <- rot_ssb$Z_rot_mean %>%
  reshape2::melt() %>%
  rename(time = Var1, model_id = Var2, loading = value) %>%
  reshape2::dcast(time ~ model_id, value.var = "loading")
```

Now we can apply any of our clustering algorithms to these loadings:

```{r}
factoextra::fviz_nbclust(dfa_loadings_fbar, cluster::pam, method = "silhouette")
factoextra::fviz_nbclust(dfa_loadings_ssb, cluster::pam, method = "silhouette")

factoextra::fviz_nbclust(dfa_loadings_fbar, cluster::pam, method = "gap_stat")
factoextra::fviz_nbclust(dfa_loadings_ssb, cluster::pam, method = "gap_stat")

detect_clusters <- cluster::clusGap(dfa_loadings_fbar, cluster::pam, K.max = 10)
.k <- cluster::maxSE(detect_clusters$Tab[,"gap"], detect_clusters$Tab[,"SE.sim"])
.k

detect_clusters_ssb <- cluster::clusGap(dfa_loadings_ssb, cluster::pam, K.max = 10)
.k_ssb <- cluster::maxSE(detect_clusters_ssb$Tab[,"gap"], detect_clusters_ssb$Tab[,"SE.sim"])
.k_ssb

m_cluster_dfa_fbar <- cluster::pam(dfa_loadings_fbar, k = .k)
plot_clusters(m_cluster_dfa_fbar,
  data = dfa_loadings_fbar,
  colour_vector = as.factor(m_cluster_dfa_fbar2$clustering),
  colour_label = "Cluster"
)

m_cluster_dfa_ssb <- cluster::pam(dfa_loadings_ssb, k = .k_ssb)
plot_clusters(m_cluster_dfa_ssb,
  data = dfa_loadings_ssb,
  colour_vector = as.factor(m_cluster_dfa_ssb2$clustering),
  colour_label = "Cluster"
)
```


```{r}
look_up <- data.frame(model_id = unique(params$timeseries_data$model_id),
  cluster = m_cluster_dfa_fbar$clustering)

params$timeseries_data %>%
  left_join(look_up) %>%
  ggplot() + geom_line(aes(
  x = year, y = fbar,
  colour = as.factor(cluster), group = model_id
)) +
  scale_colour_viridis_d()

look_up <- data.frame(model_id = unique(params$timeseries_data$model_id),
  cluster = m_cluster_dfa_ssb$clustering)

params$timeseries_data %>%
  left_join(look_up) %>%
  ggplot() + geom_line(aes(
  x = year, y = ssb,
  colour = as.factor(cluster), group = model_id
)) +
  scale_colour_viridis_d()
```
