---
title: "Dynamic Factor Analysis ensemble clustering example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ensemble-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clusterthat)
library(ggplot2)
library(dplyr)
library(bayesdfa)
theme_set(theme_minimal())
```

# An example of clustering with Bayesian Dynamic Factor Analysis

Dynamic Factor Analysis (DFA) is a timeseries dimension reduction technique that estimates a shared set of trends and loading factors on those trends for the various timeseries. bayesdfa fits Bayesian DFA via a model written in Stan. Here we will demonstrate an example with the bayesdfa package.

<https://CRAN.R-project.org/package=bayesdfa>
<http://rjournal.github.io/archive/2019/RJ-2019-007/index.html>

```{r dfa-fit}
ggplot(haddock_ts) + geom_line(aes(
  x = year, y = fbar,
  colour = model_id, group = model_id
)) +
  scale_colour_viridis_c()
```

First we need to reshape our data for fitting:

```{r}
df <- haddock_ts %>%
  reshape2::dcast(year ~ model_id, value.var = "fbar") %>%
  dplyr::select(-year)
head(df)
```

The columns are our various models and the rows represent years.

We will fit a model with 1 underlying trend for speed. We will only use one chain so that this runs quickly but you will likely want to use at least 4 and probably 1000 or greater iterations.

```{r, warning=FALSE, results="hide"}
m2 <- fit_dfa(y = df, num_trends = 1, chains = 1, iter = 300, data_shape = "long")
```

We will apply a varimax rotation to describe as much of the variation in the initial trend and then plot the underlying trends:

```{r}
rot2 <- rotate_trends(m2)
plot_trends(rot2)
```

We can also plot the loadings:

```{r, warnings = FALSE}
g <- plot_loadings(rot2)

# remove colour if desired:
g$layers[[1]] <- NULL
g + geom_violin(colour = NA, fill = "grey30") +
  coord_flip(ylim = c(-1, 1.5)) # demonstrating how to limit the axis
```

We can now use the loadings in a clustering analysis.

The following are our rotated loadings. We will reshape them into a data frame.

```{r}
head(rot2$Z_rot_mean)

dfa_loadings <- rot2$Z_rot_mean %>%
  reshape2::melt() %>%
  rename(time = Var1, model_id = Var2, loading = value) %>%
  reshape2::dcast(time ~ model_id, value.var = "loading")
```

Now we can apply any of our clustering algorithms to these loadings:

```{r}
factoextra::fviz_nbclust(dfa_loadings, cluster::pam, method = "wss")
factoextra::fviz_nbclust(dfa_loadings, cluster::pam, method = "silhouette")
factoextra::fviz_nbclust(dfa_loadings, cluster::pam, method = "gap_stat")

m_cluster_dfa <- cluster::pam(dfa_loadings, k = 2L)
plot_clusters(m_cluster_dfa,
  data = dfa_loadings,
  colour_vector = as.factor(m_cluster_dfa$clustering),
  colour_label = "Cluster"
)
```
